\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-12 analysis/manuscript.tex
refs

@@

This study combines data from multiple publicly available data sources
as diagrammatically shown in Figure \ref{fig:data_workflow}. Initially,
transformative agreement data retrieved from the cOAlition S Journal
Checker Tool provided information about journal portfolios and
participating institutions. After identification of hybrid journals,
Crossref served as the primary data source for article-level metadata
including Creative Commons (CC) license information to determine open
access availability on publisher websites. To identify open access
articles published through transformative agreements, first author
affiliations from OpenAlex were subsequently linked to eligible
institutions according to the transformative agreement data.

\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{data_collection_workflow}

@@

\caption{Data collection workflow to obtain article-level transformative agreement data}\label{fig:data_workflow}
\end{figure}

@@

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool.\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
The dump consisted of multiple online Google spreadsheets where each
data file represented one agreement listed in the ESAC Transformative
Agreement Registry.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}
From the retrieved spreadsheet files, journals and institutions involved
per agreement were obtained.

A limitation of using the Journal Checker Tool to study the development
of transformative agreements over time is that it only provided data
about active transformative agreements. As a result, expired
transformative agreements were constantly removed from the publicly
available data dump. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included. Overall, the
four combined Transformative Agreement Data dumps used in this study
contained 729 out of 869 agreements listed in the ESAC registry by
December 2023.

The Transformative Agreement Data dumps link agreements to journals
represented by journal names and ISSN. After mapping ISSN variants to
the corresponding linking ISSN (ISSN-L) as provided by the ISSN
International Centre, journals were associated to publishers using the
ESAC ID, a unique identifier for transformative agreements in the ESAC
Transformative Agreement Registry. Because transformative agreements can
include both full open access and hybrid journals, the data were
complemented with information about a journal's open access status using
multiple sources: the Directory of Open Access Journals (DOAJ)
downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended the journal
matching to exclude full open access journals from transformative
agreements. In total, 3,439 full open access journals were excluded
based on ISSN matching. The overlap between the three data sources was
72\%. The Gold OA journals dataset alone added 176 journals, while the
DOAJ comprised 10 full open access journals not listed in either of the
other two sources. These full open access journals were mostly launched
in 2022.

@@

After identifying hybrid journals included in transformative agreements,
article metadata was retrieved from Crossref\textless!. November 2023
database snapshot for the five-year period 2018 to 2022. Because
Crossref metadata lacked information to distinguish between original
research articles including review and other types of journal content,
which are often not covered by transformative agreements (Borrego et
al., 2020), only articles published in regular issues indicated by
non-numeric pagination were included. Furthermore, an expanded version
of Unpaywall's paratext recognition approach was applied to exclude
non-scholarly journal content such as table of contents.

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017). Comparing Crossref license coverage with OpenAlex,
which re-uses open access evidence from Unpaywall, not all publisher
provide such license information a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), demonstrates the ongoing challenges to identify
open access articles in hybrid journals (Jahn et al., 2021;
Martín-Martín et al., 2018). Overall, 742,369 articles with CC license
were retrieved using Crossref, while 950,260 articles were tagged as
``hybrid'' in OpenAlex. The biggest differences concerned articles
published between 2018 and 2020. In 2022, however, Crossref and OpenAlex
open access numbers only differ slightly (249,511 records using Crossref
vs.~255,344 in OpenAlex). Upon inspection, notable difference could be
furthermore observed among some publishers that presumably did not
provide CC license information to Crossref including AIP Publishing,
American Physiological Society and Emerald. Crossref license metadata
was more complete with regard to articles from the publisher Wiley and
American Chemical Society. Finally, inconsistent open access status
information in previous OpenAlex versions was observed (Jahn et al.,
2023). According to the OpenAlex release notes, fixing this issue was
still ongoing, which might also explain this discrepancy.

@@

Author metadata are crucial for the planning and evaluation of
transformative agreements (Geschuhn \& Stone, 2017; Schimmer et al.,
2015), in particular information about corresponding authors who are
generally responsible to arrange open access publication (Borrego et
al., 2020). Here, affiliations were retrieved from OpenAlex. However,
because of low coverage, this study focused on first authors and their
affiliations. First authors typically contribute most to a paper and are
often considered lead author research papers (Larivière et al., 2016).
Overall, around 90\% of studied articles had first author affiliation
metadata in OpenAlex, whereas the coverage of articles with
corresponding author information was around 54\%.

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID. The matching also
took into account the duration of an agreement according to the ESAC
registry. Upon inspection, Transformative Agreement Data did not cover
associated institutions comprehensively like university hospitals or
institutes of large research organisations like the Max Planck Society.
To improve the matching, Transformative Agreement Data was automatically
enriched with associated institutions using OpenAlex's insitution data.

In total, the compiled data set consists of 8,922,146 articles published
in 12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}. Hybrid journals included in transformative
agreements represented 40\% of total global output over the same time
period according to Crossref, while full open access journals accounted
for 0\% of total article volume.

\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{fig/method_fig-1}

@@

Throughout this mostly automated data gathering and analysis process, we
used tools from the Tidyverse (Wickham et al., 2019) for the R
programming language (R Core Team, 2020). The resulting data is openly
available through an R package, hoaddata. Following Marwick et al.
(2018), hoaddata contains not only the datasets used in the data
analysis. It also includes codeused to compile the data by connecting it
the cloud-based Google Big Query data warehouse, where the big scholarly
data from Crossref, OpenAlex and Unpaywall were imported. To increase
the computational reproducibility, the R package was automatically built
using GitHub Actions, a continuous integration service.

@@

\begin{figure}

{\centering \includegraphics[width=0.99\linewidth]{fig/unnamed-chunk-2-1}

@@

\caption{Relative growth of open access in hybrid journals through transformative agreements between 2018 and 2022 per publication year  according to the total number of articles indexed in Crossref. The blue areas represent open access through transformative agreements, as indicated by the affiliation with institutions that had a corresponding transformative agreement in place at the time of publication according to the first author affiliation data in OpenAlex matched with journal and institutional data from cOAlition S Transformative Agreements Public Data. Grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. Horizontal lines correspond to the lower quartile, median, and upper quartile, with whiskers extending to ±1.5 of interquartile range. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available at the time of this study.}\label{fig:unnamed-chunk-2}
\end{figure}

Figure 2A shows a moderate growth in the proportion of open access
articles in hybrid journals, comparing the overall open access uptake
and the impact of transformative agreements on this trend. Over the
five-years period from 2018 to 2022, open access increased from 4.3\% (n
= 65,486) to 15\% (n = 249,511). At the same time, the total article
volume of the investigated journals grew from 1,528,051 in 2018 to
1,676,928 in 2022.

Figure 2B highlights that the majority of open access articles in hybrid
journals were made available through transformative agreements in 2021
and 2022, contributing 58\% of the total open access article volume in
2022. However, there was also a notable growth in open access provision
through individual publication fees, which increased from 4.1\% (n =
62,625) in 2018 to 6.3\% (n = 105,896). This suggests that publishers
were able to gain equally from individual and institutional open access
publishing options.

Figure 2C depicts the substantial variations among the hybrid journals
included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure 2A, the
farther stretch of upper quartiles and whiskers over the years
illustrates that an increasing number of journals published an
above-average proportion of open access articles. In 2022, 25\% of
hybrid journals (n = 2,576) had an open access uptake of 29\%, and 6.6\%
of journals (n = 744) provided the majority of their articles under a
Creative Commons license in the same year. These journals were, on
average, smaller (M = 75, SD = 186) than those with an open access share
below 50\% (M = 164, SD = 347). Notable exception of large journals with
an above-average open access proportion were Physical Review D, a
high-energy physics journal covered by the SCOAP3 consortium that
provided open access to 2,341 out of 4,074 articles in 2022, Astronomy
and Physics (1,396 out of 2,230 articles in 2022 were open access),
which shifted to a subscribe to open business model for all accepted
articles as of April 2022, the Journal of Fluid Mechanics (577 of 1,077
articles in 2022 were open access) and Bioinformatics ( out of articles
in 2022 were open access), which flipped to full open access as of
January 2023.

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure 2D). Examples of such journals include those with a
scope on specific countries or regions, where also transformative
agreements were implemented. For instance, in 2022, the Germany-based
journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für
Politikwissenschaft, as well as the Scandinavian Political Studies
adressing the Nordic countries, achieved an overall open access uptake
of more than 90\% just through transformative agreements. Despite the
rise in transformative agreements, it is worth noting that other means
of publishing open access in hybrid journals remained common. In total,
9,153 journals published open access articles from authors affiliated
with institutions without transformative agreements in place, while
8,780 journals published at least one open access article through a
transformative agreement in the same year.

@@

\caption{\label{tab:unnamed-chunk-5}Hybrid open access through transformative agreements market shares 2018-2022}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrllrllrllrll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Hybrid journals} & \multicolumn{3}{c}{Articles} & \multicolumn{3}{c}{OA articles} & \multicolumn{3}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
Publisher & Total & \% & C\% & Total & \% & C\% & Total & \% & C\% & Total & \% & C\%\\
\midrule
Elsevier & 1,936 & 17 & 17 & 2,770,826 & 33.8 & 33.8 & 172,723 & 22.9 & 22.9 & 60,440 & 18.3 & 18.3\\
Springer Nature & 2,274 & 20 & 37 & 1,330,430 & 16.2 & 50 & 175,432 & 23.3 & 46.2 & 100,008 & 30.3 & 48.6\\
Wiley & 1,410 & 12.4 & 49.4 & 1,043,052 & 12.7 & 62.7 & 152,723 & 20.3 & 66.5 & 83,443 & 25.3 & 73.9\\
Other & 5,767 & 50.6 & 100 & 3,061,337 & 37.3 & 100 & 252,523 & 33.5 & 100 & 86,294 & 26.1 & 100\\
\bottomrule
\end{tabular}
\end{table}

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table ). Together, they published 500,878 open access
articles, representing 66\% of the open access articles in hybrid
journals. Elsevier, Springer Nature, and Wiley made 243,891 articles
open access in hybrid journals through transformative agreements,
resulting in an even larger market share of 74\%. However, there are
differences among the three large publishers. Although Elsevier
published the largest volume of articles (n = 2,770,826, 34\%), it
published a relatively low number of open access articles, including
those that can be associated with transformative agreements. In
contrast, Springer Nature and Wiley provided open access to a larger
proportion of their articles (13\% of Springer Nature articles and 15\%
of Wiley articles were open acccess), leading to higher open access
market shares (23\% Springer Nature resp. 23\% Wiley). This difference
between Elsevier on the one hand and Springer Nature and Wiley on the
other can be attributed to transformative agreements, as the latter made
the majority of their open access articles available through such deals
(Springer Nature 57\% resp. Wiley 55\%).

\begin{center}\includegraphics[width=0.99\linewidth]{fig/unnamed-chunk-6-1} \end{center}

Figure 2 takes a closer look into the growth of hybrid open access
across publishers by year with a focus on open articles enabled by
transformative agreements. Although all publishers show a general
long-term trend towards transformative agreements, Figure 2A and 2B
indicate that, in particular, Wiley's has experienced a substantial
increase in its open access share from 5.9\% (n = 11,628) in 2018 to
26\% (n = 53,503) in 2022, rperesenting an 4.5-fold increase. In
contrast, Elsevier's hybrid journals deonstrated a more modest increase,
from 3.3\% (n = 16,872) in 2018 to 10\% (n = 60,821) in 2022, which is a
relatively low open access share compared to the general trend. In 2018,
Springer Nature had the largest open access proportion among the three
publishers of 8.4\% (n = 19,701), but experienced a relatively slower
growth, resulting in 18\% (n = 52,616) of articles being open access in
Springer Nature hybrid journals in 2022.

@@

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure 2). While no substantial
differences between open access enabled through transformative
agreements and other revenue source could observed across Elsevier
journals, the distribution of open access across Springer Nature and
Wiley hybrid journals indicates that the growth is not limited to a few
journals, but extends across the portfolio. In particular, Wiley's upper
quantile, which represents the top 25\% of journals in terms of the
proportion of open access articles from transformative agreements,
increased markedly from 13\% in 2020 to 31\% in 2022. At the same time,
the median proportion grew from 7.5\% to 19\%. It is interesting to note
that a small but increasing number of journals from these two publishers
are providing open access to the majority of articles through
transformative agreements. Wiley recorded 68 and Springer Nature 102
hybrid journals with an open access share above 50\% that could be
solely attributed to transformative agreements. Upon inspection, these
journals were mainly society or local language journals with a small
yearly article volume.

@@

Table presents a high-level overview of hybrid open access by AJCS
subject area using fractionalised counting to account for journals
belonging to more than one category. Between 2018 and 2022, most hybrid
journals with at least one open articles could be attributed to the
social sciences including the humanities. However, these journals
published the fewest number of articles, whereas physical sciences
journals recorded most articles, followed by the health sciences and the
life sciences. In terms of open access, physical sciences journals
accounted for more than one third of articles published in the
five-years period, followed by the health science, the social sciences
and the life sciences.

@@

\begin{center}\includegraphics[width=0.99\linewidth]{fig/country_patch-1} \end{center}

>>>>

For over two decades, hybrid open access journal publishing, which makes
some articles freely available under an open license while others remain
behind a paywall, has been discussed as a model for transitioning the
subscription system to full open access (Prosser, 2003). The idea was
that when journals publish more and more open access articles, they
could reduce revenues from subscriptions, while libraries and funders
could change their funding models and shift expenditures from
subscription to open access. However, initial approaches, mainly based
on publication fees, also called article processing charges (APCs), did
not contribute substantially to an large open access uptake. In 2012 ddd
- at the same time

Tthe introduction of transformative agreements, a collective term for
agreements starting from mid-2015, has resulted in a new push for hybrid
open access. In particular, the Max Planck Society's Berlin 2020
initiative and plan has advocated these types of agreements as a
transitional means to progress with open access. As a result, more than
800 transformative agreements have been tracked so far, which have been
instrumental in funding activities and controversies. today, all
publsiehrs offer such agreements

Advocates of these agreements argue that they not only overcome the
issue of double-dipping but also help to shift library spending from
subscriptions to publications, streamline expenditures, and offer more
value than big deals, which focus on large publishers. A few agreements
can impact an institution's open access share, resulting in a
significant proportion of open access content. increase transparency

However, opponents of transformative agreements perceive continued
reliance on big deals as problematic. They argue that the focus on large
publishers perpetuates the business relationships that led to the serial
crisis and lacks evidence that this leads to a reduction of In addition,
concerns over inequality abound. could not benn mass removal

Effects can be seen, mixed consequences. some countries who have
introduced this fundign effects netherlands, sweden. no in germany,
elsevier deal to price reduction and first steps towards mroe privacy.
however, main goals associated woth trnasofrtiomative, the transition of
the journal landscape not achivned, publsiehrs were not able to meet
growtht raget, sweden did not see the flipping and has

This study contributes

@@

This study combines data from multiple publicly available data sources
as diagrammatically shown in Figure \ref{fig:data_workflow}. Initially,
transformative agreement data retrieved from the cOAlition S Journal
Checker Tool provided information about journal portfolios and
participating institutions. After identification of hybrid journals by
excluding full open access journals, Crossref served as the primary data
source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. To determine open access articles published through
transformative agreements, first author affiliations from OpenAlex
(Priem et al., 2022) were subsequently linked to eligible institutions
according to the transformative agreement data.

\begin{figure}[ht!]

{\centering \includegraphics[width=0.99\linewidth,]{data_collection_workflow}

@@

\caption{Data collection workflow}\label{fig:data_workflow}
\end{figure}

@@

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool.\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
The dump consists of multiple online Google spreadsheets where each data
file represents one agreement listed in the ESAC Transformative
Agreement Registry.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}
From the retrieved spreadsheet files, journals and institutions involved
per agreement were obtained.

A limitation of using the Journal Checker Tool and its underlying
publicly available data dump to study the development of transformative
agreements over time is that expired transformative agreements are
constantly removed. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included, representing
the majority of transformative agreements. Overall, the four combined
Transformative Agreement Data dumps used in this study contained 729 out
of 869 agreements listed in the ESAC registry by December 2023.

The Transformative Agreement Data dumps link agreements to journals
represented by journal names and ISSN. After mapping ISSN variants to
the corresponding linking ISSN (ISSN-L) as provided by the ISSN
International Centre, journals were associated to publishers using the
ESAC ID, a unique identifier for transformative agreements in the ESAC
Transformative Agreement Registry. Furthermore, journal subjects
according to the All Science Journal Classification code (ASJC) were
added from the Scopus journal source list as of August 2023.

Because transformative agreements can include both full open access and
hybrid journals, the data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended the journal
matching to exclude full open access journals from transformative
agreements. In total, 3,439 full open access journals were excluded
based on ISSN matching. The overlap between the three data sources was
72\%. The Gold OA journals dataset alone added 176 journals, while the
DOAJ comprised 10 full open access journals not listed in either of the
other two sources. These full open access journals were mostly launched
in 2022.

@@

After identifying hybrid journals included in transformative agreements,
article metadata was retrieved from the Crossref November 2023 database
snapshot for the five-year period 2018 to 2022 according to the issued
date, representing the earliest known publication date. Because Crossref
metadata lacked information to distinguish between original research
articles including review and other types of journal content, which are
often not covered by transformative agreements (Borrego et al., 2020),
only articles published in regular issues indicated by non-numeric
pagination were included. Furthermore, an expanded version of
Unpaywall's paratext recognition approach was applied to exclude
non-scholarly journal content such as table of contents.

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017). Comparing Crossref license coverage with OpenAlex,
which re-uses open access evidence from Unpaywall, a widely used open
access discovery service that also parses journal websites for open
content licenses (Piwowar et al., 2018), highlighted ongoing challenges
to identify hybrid open access (Butler et al., 2023; Jahn et al., 2021;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society and Emerald. Crossref license metadata
was more complete with regard to articles from the publisher Wiley and
American Chemical Society. Finally, inconsistent open access status
information in previous OpenAlex versions was observed (Jahn et al.,
2023). According to the OpenAlex release notes, fixing this issue was
still ongoing, which might also explain this discrepancy.

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements, because they are
considered to be responsible to arrange open access publication (Borrego
et al., 2020; Geschuhn \& Stone, 2017; Schimmer et al., 2015). Here,
country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on
first authors and their affiliations instead. First authors typically
contribute most to a paper and are often considered lead author research
papers (Larivière et al., 2016), and can be therefore assumed as a proxy
to measure to open access payments (Zhang et al., 2022). Overall, around
90\% of studied articles had first author affiliation metadata in
OpenAlex, whereas the coverage of articles with corresponding author
information was around 54\%.

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID. The matching also
took into account the duration of an agreement according to the ESAC
registry. Upon inspection, Transformative Agreement Data did not cover
associated institutions comprehensively like university hospitals or
institutes of large research organisations like the Max Planck Society.
To improve the matching, Transformative Agreement Data was automatically
enriched with associated organisations using OpenAlex's institution
entity.

In total, the compiled data set consists of 8,922,146 articles published
in 12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals included in transformative
agreements represented 40\% of total global output over the same time
period according to Crossref, while full open access journals accounted
for 35\% of total article volume.

\begin{figure}[ht!]

{\centering \includegraphics[width=0.99\linewidth,]{fig/method_fig-1}

@@

Throughout this mostly automated data gathering and analysis process, we
used tools from the Tidyverse (Wickham et al., 2019) for the R
programming language (R Core Team, 2020). The resulting data is openly
available through an R package, hoaddata. Following Marwick et al.
(2018), hoaddata contains not only the datasets used in the data
analysis. It also includes code used to compile the data by connecting
it the cloud-based Google Big Query data warehouse, where the big
scholarly data from Crossref, OpenAlex and Unpaywall were imported. To
increase the computational reproducibility, the R package was
automatically built using GitHub Actions, a continuous integration
service.

@@

\begin{figure}[ht!]

{\centering \includegraphics[width=0.99\linewidth,]{fig/results_overview-1}

@@

\caption{Relative growth of open access in hybrid journals in transformative agreements between 2018 and 2022 per publication year. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established (according to matching OpenAlex first author affiliations matched with cOAlition S transformative agreement data). (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:results_overview}
\end{figure}

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). At the same
time, the total article volume of the investigated journals grew from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of open
access articles in hybrid journals were made available through
transformative agreements in 2021 and 2022, contributing 58\% of the
total open access article volume in 2022. However, there was also a
notable growth in open access provision through individual publication
fees, which increased from 4.1\% (n = 62,625) in 2018 to 6.3\% (n =
105,896). This suggests that publishers were able to gain equally from
individual and institutional open access publishing options.

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. These journals were, on average, smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).
Notable exception of large journals with an above-average open access
proportion were Physical Review D, a high-energy physics journal covered
by the SCOAP3 consortium that provided open access to 2,341 out of 4,074
articles in 2022, Astronomy and Physics (1,396 out of 2,230 articles in
2022 were open access), which shifted to a subscribe to open business
model for all accepted articles as of April 2022, and the Journal of
Fluid Mechanics (577 of 1,077 articles in 2022 were open access).

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure \ref{fig:results_overview}D). Examples of such
journals include those with a scope on specific countries or regions,
where also transformative agreements were implemented. For instance, in
2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft
and Zeitschrift für Politikwissenschaft, as well as the Scandinavian
Political Studies adressing the Nordic countries, achieved an overall
open access uptake of more than 90\% just through transformative
agreements. Despite the rise in transformative agreements, it is worth
noting that other means of publishing open access in hybrid journals
remained common. In total, 9,153 journals published open access articles
from authors affiliated with institutions without transformative
agreements in place, while 8,780 journals published at least one open
access article through a transformative agreement in the same year.

@@

\caption{\label{tab:publisher_league_table}Hybrid open access through transformative agreements market shares 2018-2022}
\centering
\begin{tabular}[t]{lrlrlrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hybrid journals} & \multicolumn{2}{c}{Articles} & \multicolumn{2}{c}{OA articles} & \multicolumn{2}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Publisher & Total & \% & Total & \% & Total & \% & Total & \%\\
\midrule
Elsevier & 1,936 & 17 & 2,770,826 & 33.8 & 172,723 & 22.9 & 60,440 & 18.3\\
Springer Nature & 2,274 & 20 & 1,330,430 & 16.2 & 175,432 & 23.3 & 100,008 & 30.3\\
Wiley & 1,410 & 12.4 & 1,043,052 & 12.7 & 152,723 & 20.3 & 83,443 & 25.3\\
Other & 5,767 & 50.6 & 3,061,337 & 37.3 & 252,523 & 33.5 & 86,294 & 26.1\\
\bottomrule
\end{tabular}
\end{table}

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 open access articles, representing 66\% of the open
access articles in hybrid journals. Elsevier, Springer Nature, and Wiley
made 243,891 articles open access in hybrid journals through
transformative agreements, resulting in an even larger market share of
74\%. However, there are differences among the three large publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it published a relatively low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

\begin{figure}[ht!]

{\centering \includegraphics[width=0.99\linewidth,]{fig/publisher_figure-1}

\caption{Developement of open access in hybrid journals included in transformative agreements between 2018 and 2022 by publishers. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established (according to matching OpenAlex first author affiliations matched with cOAlition S transformative agreement data). (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:publisher_figure}
\end{figure}

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year with a focus on open
articles enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that, in particular, Wiley's
has experienced a substantial increase in its open access share from
5.9\% (n = 11,628) in 2018 to 26\% (n = 53,503) in 2022, representing an
4.5-fold increase. In contrast, Elsevier's hybrid journals demonstrated
a more modest increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n =
60,821) in 2022, which is a relatively low open access share compared to
the general trend. In 2018, Springer Nature had the largest open access
proportion among the three publishers of 8.4\% (n = 19,701), but
experienced a relatively slower growth, resulting in 18\% (n = 52,616)
of articles being open access in Springer Nature hybrid journals in
2022.

@@

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure \ref{fig:publisher_figure}C).
While no substantial differences between open access enabled through
transformative agreements and other revenue source could observed across
Elsevier journals, the distribution of open access across Springer
Nature and Wiley hybrid journals indicates that the growth is not
limited to a few journals, but extends across the portfolio. In
particular, Wiley's upper quantile, which represents the top 25\% of
journals in terms of the proportion of open access articles from
transformative agreements, increased markedly from 13\% in 2020 to 31\%
in 2022. At the same time, the median proportion grew from 7.5\% to
19\%. It is interesting to note that a small but increasing number of
journals from these two publishers are providing open access to the
majority of articles through transformative agreements. Wiley recorded
68 and Springer Nature 102 hybrid journals with an open access share
above 50\% that could be solely attributed to transformative agreements.
Upon inspection, these journals were mainly society or local language
journals with a small yearly article volume.

@@

Table \ref{tab:subject_summary_table} presents a high-level overview of
hybrid open access by AJCS subject area using fractionalised counting to
account for journals belonging to more than one category. Between 2018
and 2022, most hybrid journals with at least one open articles could be
attributed to the social sciences including the humanities. However,
these journals published the fewest number of articles, whereas physical
sciences journals recorded most articles, followed by the health
sciences and the life sciences. In terms of open access, physical
sciences journals accounted for more than one third of articles
published in the five-years period, followed by the health science, the
social sciences and the life sciences.

@@

\begin{center}\includegraphics[width=0.99\linewidth,]{fig/countrypatch-1} \end{center}

Figure \ref{fig:countrypatch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa in 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals (see Supplement). While OECD
publication output in Elsevier's Physical Sciences journals declined
from 112,822 articles in 2018 to 103,766 in 2022, BRICS output increased
from 104,654 to 171,713 in the same five-year period. Furthermore, OECD
publication output in Health Science Journals and Life Science journals
stagnated across the investigated hybrid journal portfolios after a peak
in 2020.
